---
title: "Bayesian Probabilistic Selection index"
author: "José Tiago Barroso Chagas"
e-mail: "josetchagas@usp.br"
orcid: "0000-0001-6123-7413"
institute: "ESALQ/USP"
output:
  rmdformats::html_clean:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    toc: true
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

# Introduction

This workflow will help you in the Bayesian probabilistic selection index (BPSI) usage.
To apply this tool for your crop, it's recommended pay attention to each section and run carrefully the codes.
Any questions? Contact us by \url{https://github.com/saulo-chaves/ProbBreed} or e-mail.

## Pan-African Soybean Trials dataset

Cultivar recommendation is a critical stage in plant breeding programs, and selecting superior genotypes for multiple traits remains a challenge due to the genotypes × environment (G × E) interaction. 🌱 

To address this, we analyzed multi-environment trials data from Pan-African Trials, evaluating 62 soybean genotypes across 19 environments [@Araujo2025]. 🌦️ 

Our methodological approach integrates Bayesian probabilistic framework to estimate genotypes risk recommendation to each trait and select genotypes according to a multitrait ideotype [@Chagas2025]. The Bayesian probabilistic selection index (BPSI)  select the 10% genotypes with higher probability of superior performance across environments to Grain Yield, Plant Height and Number of Days to Maturity. 💡


## Bayesian probabilistic selection index (BPSI)

The Bayesian probabilistic selection index enables design ideotype by increase or decrease and weighting traits.
Along with the probabilities of superior performance across environments, it accounts for the distance between the worst genotype performance and candidate genotypes for each trait.



$$
BPSI_i = \sum_{m=1}^{t} \frac{\gamma_{pt} -\gamma_{it} }{(1/\lambda_t)}
$$

where $\gamma_p$ is the probability of superior performance of the worst genotype for the trait $m$, $\gamma$ is the probability of superior performance of genotype  $i$ for trait $m$,
$t$ is the total number of traits evaluated, $\left(m = 1, 2, ..., t \right)$, and $\lambda$ is the weight for each trait $t$.




# BPSI example at soybean dataset

In order to provide a step-by-step guide, we applied 


## Bayesian Model

The Bayesian model are available at ProbBreed package [@Chaves2024]. We used the entry mean model.

$$
y_{jk} = \mu + l_k + g_j  + \varepsilon_{jk}
$$
where the $y_{jk}$  is the phenotypic record of the  genotype $j^{th}$ in the  $k^{th}$  location, $\mu$   is the intercept, $l_k$   is the main effects of the  location,  $g_j$  is the main effect of the  genotype,  and $\epsilon_{jk}$ is the residual effect.

```{r}
#| echo: true
#| warning: false
#| message: false


library(ProbBreed)

met_df=read.csv("https://raw.githubusercontent.com/tiagobchagas/BPSI/refs/heads/main/Data/blues_long.csv",header=T)

head(met_df)
mod = bayes_met(data = met_df,
                gen = "gen",
                loc = "env",
                repl = NULL,
                trait = "PH",
                reg = NULL,
                year = NULL,
                res.het = T,
                iter =40, cores = 4, chain = 4) #recommended run at least 4k iterations


mod2 = bayes_met(data = met_df,
                 gen = "gen",
                 loc = "env",
                 repl = NULL,
                 trait = "GY",
                 reg = NULL,
                 year = NULL,
                 res.het = T,
                 iter = 40, cores = 4, chain = 4) #recommended run at least 4k iterations

mod3 = bayes_met(data = met_df,
                 gen = "gen",
                 loc = "env",
                 repl =  NULL,
                 trait = "NDM",
                 reg = NULL,
                 year = NULL,
                 res.het = T,
                 iter = 40, cores = 4, chain = 4) #recommended run at least 4k iterations
```

## Extracting the probability of superior performance

```{r}
#| echo: true
#| warning: false
#| message: false


models=list(mod,mod2,mod3)
names(models) <- c("PH","GY","NDM")
inc=c(FALSE,TRUE,FALSE)
names(inc) <- names(models)

results <- lapply(names(models), function(model_name) {
  x <- models[[model_name]]  # actual model object

  outs <- extr_outs(model = x,
                    probs = c(0.05, 0.95),
                    verbose = TRUE)

  a <- prob_sup(extr = outs,
                int = .2,
                increase = inc[[model_name]],  # ← now model_name is a character!
                save.df = FALSE,
                verbose = TRUE)

  return(a)
})
names(results) <- names(models)
```

## Running the BPSI

```{r}
#| echo: true

source("Data/bpsi.R")



bpsi=BPSI(problist=results,
          increase = c(FALSE,TRUE,FALSE),
          int = 0.1,
          lambda =c(1,2,1),
          save.df = F)
```

## BPSI Ranks

The selected genotypes (`r bpsi$BPSI[which(bpsi$BPSI$sel=="Selected"),"gen"]`) have superior performance for the three traits evaluated Grain Yield (GY), Plant Height (PH) and Number of Days to Maturity (NDM).

```{r}
#| echo: true
#| fig-cap: "Rank of probability of superior performance of soybean Pan-African Trials across locations in Kenya"

plot(bpsi,category = "Ranks")
```

## BPSI plot
The selected genotypes (`r bpsi$BPSI[which(bpsi$BPSI$sel=="Selected"),"gen"]`) have minor risk of cultivar recomendation to the multitrait ideotype across the Kenya environments.
```{r}
#| echo: true
#| fig-cap: "Bayesian probability selection index of soybean genotypes from Pan-African Trials across locations in Kenya"

plot(bpsi,category = "BPSI")
```


## Probabilities of superior performance 
The probability os superior performance across environments show the risk recommendation of the soybean genotypes being in the top 20% to Grain Yield (GY) and the bottom 20% for Plant Height (PH) and Number of Days to Maturity (NDM). 

```{r}
df=bpsi$df
df |> kableExtra::kable()
```

## Genotypes selected
The genotypes by BPSI selected were `r bpsi$BPSI[which(bpsi$BPSI$sel=="Selected"),"gen"]`. It have the greater distance to the worst genotype to Grain Yield (GY), Plant Height (PH) and Number of Days to Maturity (NDM). The genotype selected are the 10% multitrait top performance across enviroments.
```{r}
df=bpsi$BPSI

df |> kableExtra::kable()
```

